---
title: Perimeter of Ignorance
author: ''
date: '2018-02-18'
tags:
  - Quotes
  - Visualization
  - ggplot2
slug: perimeter-of-ignorance
lastmod: '2019-08-26T19:21:48-04:00'
featured: no
summary: "A quote that is easily visualized, because a picture is worth a thousand ~~words~~ lines of code"
image:
  caption: 'As _A_ increases, so does _C_'
  placement: 3
  preview_only: no
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

> _"As the area of our knowledge increases, so too, does the perimeter of our ignorance"_  — Neil DeGrasse Tyson

One of the many quotables from one of the great science communicators.  A key reason in selecting this quote is due to its graphical representation.  

The visualization above shows how increasing _area_ ($A$) leads to an increase in _perimeter_ ($C$) by virtue of the increase in radius ($r$) which is the key component of both equations.  

_**NB**:_ Some who are particularly astute (or anal) might correctly highlight that paralleling _ignorance_ to "perimeter" is slightly incorrect: it's the area of the surrounding space - exclusive of the circle - that represents ignorance, and this area certainly decreases as the circle gets bigger .. but nobody likes a keener  

Here's the R code to create the visualization above using `ggplot2`, `ggforce` (for an easy circle _geom_), and `viridis` for colour pallette:  

```{r circle, eval=FALSE}
library("ggplot2")
library("ggforce")
library("viridis")

#data.frame that provides all the requisite data for plotting
circle <- data.frame(x = c(1, 7),
                     y = rep(2, 2),
                     r = c(3, 2),
                     type = ordered(c("More", "Less")))
#adding `area` (A) and `circumference` (C)
circle <- circle %>% 
        mutate(A = round(pi * (circle$r)^2, 1),
               C = round(2 * pi * (circle$r), 1))

#the vis
ggplot(data = circle) + 
        # the plot
        geom_circle(aes(x0 = x, y0 = y, r = r, fill = type), 
                    colour = "slategrey") + 
        annotate("text", x = circle$x, y = circle$y + 0.25, 
                 label = paste0("A ≈ ", circle$A, "u²"),
                 hjust = 0, 
                 colour = "slategrey") +
        annotate("text", x = circle$x, y = circle$y + circle$r + 0.25, 
                 label = paste0("C ≈ ", circle$C, "u"),
                 colour = "slategrey") +
        annotate("text", x = -1.8, y = -1.25, 
                 label = "Area==pi*r^2~~and~~Circumference==2*pi*r", 
                 parse = T,
                 hjust = 0, 
                 colour = "slategrey") +
        # a colour pallette from viridis
        scale_fill_viridis(discrete = T, option = "C") +
        coord_fixed() +
        # vis labels
        labs(x = "", y = "", fill = "Knowledge") +
        theme_bw() + 
        theme(legend.position = "bottom",
              axis.text = element_blank(),
              axis.ticks = element_blank(),
              panel.grid = element_blank(),
              plot.background = element_blank(),
              panel.border = element_blank())
```